Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2024-07-15 analysis/paper/paper.qmd
work on the intro section

---
title: "An Environmental Perspective on European Colonization in Northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute:
        - AS
institute:
  - AS:
      name: Academia Sinica
      address: 
  - IHP: 
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

# Introduction

European expansion since the 15th century marked the beginning of an era characterized by exploration, colonization, and global trade. Due to purposes and situations, the mode, pace, and degree of colonial control varied greatly across regions [@Jordan2009]. Understanding the processes of different colonial encounters has become a significant issue in archaeology to comprehend the dynamic outcomes of European colonization worldwide. In Asia, European colonization was primarily driven by the pursuit of Asian luxury goods and local natural resources for commercial profit. The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, Indigenous communities did not passively accept European colonization. Recent discussions on colonial interactions also highlight the diverse strategies employed by Indigenous communities in response to European colonization, including resistance, negotiation, incorporation, persistence, and resilience [@Torrence2003; @Silliman2005; @Mitchell2010].

While many researchers focus on European colonies, places on the periphery of colonial centers were also significantly influenced by European powers and their associated trading systems [@Acabado2017; @Trabert2017]. These peripheral regions often experienced distinct forms of interaction and impact due to their unique geographical, cultural, and economic contexts. Exploring these areas is essential for gaining a broader understanding of the varied consequences of colonial expansion. Northeastern Taiwan is a good example for such colonial setting. Although it was on the periphery of European fortresses in Northern Taiwan in the 17th century, it was intensively involved in the regional exchange network stimulated by Europeans supported by the finding of large amounts of foreign trade goods at the site [@citation]. Additionally, Northeastern Taiwan was involved in the deerskin exchange system when the VOC (Dutch East India Company) formalized deerskin exports in the mid-17th century [@Cheng2017]. The impacts of European economic demands on the local economy and environment in Northeastern Taiwan still require further investigation. Understanding these effects can provide valuable insights into the ways in which Indigenous communities adapted to and were influenced by European colonial activities.

In this paper, I will explore the Indigenous experience of European colonization by examining archaeological evidence from the Kiwulan site (1350-1950 CE), a large indigenous settlement in northeastern Taiwan [@citation]. Kiwulan experienced periods of European presence in Northern Taiwan in the 17th century, followed by Han Chinese migration in the 19th century. Archaeological evidence related to subsistence practices is particularly valuable for this investigation, as the choices made in these practices can reveal the strategies Indigenous people employed to cope with European disruptions and the associated changes in their environment. Key research questions are: What were the resource use patterns of Indigenous people during the time of European colonization? Can we identify shifts in these patterns in response to colonial pressures? The later period of Chinese migration will be also included to have a long-term view. To address these questions, I will examine the faunal and plant remains from Kiwulan, with a preliminary analysis on ceramic residues. By adopting an environmental and ecological perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

## the mode of colonization

@@

# colonial studies in Archaeology
In this research, I will explore whether and how the introduction of foreign materials and practices, new forms of production, and migrated people from Europe and Chinese contributed to the changes in Indigenous social, environmental, and ecological systems since the protohistoric period (Liu and Wang, 2017). I will focus on archeological evidence from different parts in northeastern Taiwan to compare Indigenous reactions in a pericolonial context (Acabodo, 2017). Yilan is an ideal location as a starting point because of the frequent foreign contact with Europeans and Chinese traders in the protohistoric and historic period. Also, the ecological variation across the landscape from the coastal area to mountains provides multiscalar variability for exploring the effect of colonial contact on a regional scale. As a place located in the surrounding area of the European colonial base in northern Taiwan, Yilan enables a pericolonial perspective to understand the effect of European colonization. I will use historical ecological approaches to examine the interactions of people and environment in a colonial context, and their consequences on the development of the past and contemporary culture in a long-term view (Braje, 2017). This project will help us to understand social and ecological reactions of Indigenous communities to European colonization and cross-cultural contact that has not yet been fully understood.

>>>>

---
title: "Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute:
        - AS
institute:
  - AS:
      name: Academia Sinica
      address: 
  - IHP: 
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world. Due to specific spatial and temporal contexts, the forms of European colonization and power relations varied across regions [@Jorden2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was primarily driven by the pursuit of Asian luxury goods and local natural resources for commercial profit. The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, Indigenous communities did not just passively accept European colonization. Recent discussions on colonial interactions also highlight the diverse strategies employed by Indigenous communities in response to European colonization, such as resistance, negotiation, incorporation, persistence, or resilience [@Torrence2003; @Silliman2005; @Mitchell2010].

<!-- NE Taiwan as a case study -->
While many studies focus on the center of European colonies, places on the periphery of colonial center were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Northeastern Taiwan, specifically Yilan, serves as an example of a peripheral case in the 17th century when Europeans established their colonial base in northern Taiwan. Due to its geologically challenging and relatively inaccessible location, Yilan experienced limited direct European control. However, local settlements were frequently involved in the regional exchange network stimulated by the colonial presence in the north. In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export, involving many Indigenous communities and Han Chinese. Despite some discussions on the policies and operations of the deerskin trade, direct archaeological investigations into the reactions of local Indigenous people to these commercial activities are rare. Prior to the deerskin trade, deer were already a substantial part of local subsistence. Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining archaeological evidence from the Kiwulan site (1350-1950 CE), a large Indigenous settlement in northeastern Taiwan [@citation]. Kiwulan experienced periods of European presence in northern Taiwan in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. The later period of Chinese migration will also be included to provide a long-term perspective. Archaeological evidence related to subsistence practices is particularly focused in this research, as the choices made in these practices can reveal the strategies Indigenous people employed to cope with European colonial activities and the associated changes in their environment. My research questions are: What were the resource use patterns of Indigenous people before and after the European presence? Can we identify shifts in these patterns in response to colonial economic demands? To address these questions, I will examine the faunal and plant remains from Kiwulan, alongside a preliminary analysis of ceramic residues. By adopting an environmental and ecological perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

# Historical background
As an island close to the Asian continent, Northeast Asia, and Island Southeast Asia, Taiwan has been a place with frequent cross-cultural interactions since the Neolithic period and involved in a complex global trading network in the 17th century due to the European expansion. Based on the commercial purposes, the Europeans, including the Spanish and the Dutch, built fortresses and trading posts at different parts of Taiwan.

@@

# colonial studies in Archaeology

In this research, I will explore whether and how the introduction of foreign materials and practices, new forms of production, and migrated people from Europe and Chinese contributed to the changes in Indigenous social, environmental, and ecological systems since the protohistoric period (Liu and Wang, 2017). I will focus on archeological evidence from different parts in northeastern Taiwan to compare Indigenous reactions in a pericolonial context (Acabodo, 2017). Yilan is an ideal location as a starting point because of the frequent foreign contact with Europeans and Chinese traders in the protohistoric and historic period. Also, the ecological variation across the landscape from the coastal area to mountains provides multiscalar variability for exploring the effect of colonial contact on a regional scale. As a place located in the surrounding area of the European colonial base in northern Taiwan, Yilan enables a pericolonial perspective to understand the effect of European colonization. I will use historical ecological approaches to examine the interactions of people and environment in a colonial context, and their consequences on the development of the past and contemporary culture in a long-term view (Braje, 2017). This project will help us to understand social and ecological reactions of Indigenous communities to European colonization and cross-cultural contact that has not yet been fully understood.
